Renewal and stability in populations structured by remaining years of life

### 2015-10-19 ThanoRepro/ThanoRepro.tex
.

I describe the Lotka-Euler-Leslie renewal model when structured by
thanatological age. The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines of contemporary demographic research are relatable to the renewal model, and by extention to stable population theory.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here aim to provide a broader treatment allowing for
growth in the population.
The chronological-age classified renewal model and the thanatological renewal model both describe the movement of population stocks over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox available to approach complex
temporal population structure. The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph.

\section*{A lifetable approximation of thanatological age}
%Direct information on the timing of demographic transitions with respect to
%death is available retrospectively from linked population registers or
%panel surveys with mortality follow-ups.
%Some surveys also ask respondents in various ways about their perceived future
% lifetime, which may underly those demographic transitions governed to some degree by volition. 
In general, to make a lifetable-based projective statement about the remaining
time distribution of a current stock of population, it is best to to account for population subgroups
with different mortality and to account for uncertainty about future changes in
mortality.
The present inquiry approximates thanatological age using lifetable-projected remaining lifetime,
which is intended to approximate remaining lifetime under the strong
assumptions of homogenous populations und fixed vital rates. 
%Chronological age-classified counts are redistributed over future death times
% according to a fixed mortality schedule.
%In general, a lifetable approach to transforming chronologically classified
%data into thanatologically classified data is sound if the given lifetable
%accurately describes the future mortality of the age-classified population, and
%if the chronological and thantological age patterns are independent.
In this case, each birth cohort undergoes the same
pattern of attrition. In a period setting, this means that we can distribute the
remaining lifespans of the population in each age class using the
age-conditioned lifetable.
%This is the case we explore, and so the following formulas assume a single lifetable pattern.

@@

A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
$d_a$ column of the lifetable (here with single ages):
\begin{equation}
\label{eq:dyadiscret}
f_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \ec
\end{equation}
where $L_a$ is lifetable exposure.\footnote{Equation~\eqref{eq:dyadiscret}
will incur some error depending on the method used to derive $L_a$. To remove
this error, rescale over the $a$ margin so that each distribution conditioned on
$a$ sums to 1. Working with $\ell_a$ rather than $L_a$ eliminates error and
yields similar results, but is not necessarily better.} One produces in this way a probability density function over future death times for each chronological age,
$a$. The $f(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using\footnote{An identical
definition is found in \citet{brouard1989mouvements}.}

@@

Since $1 = \int_0^\infty f(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age. For
example, we can break down a population pyramid into thanatological age classes.
Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data, assuming the
2010 period mortality schedule remains constant. When looking from the
chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this recomposition for the same 2010
US data.

>>>>

I describe the Lotka-Euler-Leslie renewal model when structured by
thanatological age. The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines of contemporary demographic research are relatable to the renewal model, and by extention to stable population theory.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here provide a broader treatment allowing for
growth in the population.
The chronological-age classified renewal model and the thanatological renewal model both describe the movement of population stocks over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox available to approach complex
temporal population structure. The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph.

\section*{A lifetable approximation of thanatological age}
%Direct information on the timing of demographic transitions with respect to
%death is available retrospectively from linked population registers or
%panel surveys with mortality follow-ups.
%Some surveys also ask respondents in various ways about their perceived future
% lifetime, which may underly those demographic transitions governed to some degree by volition. 
In general, to make a lifetable-based projective statement about the remaining
time distribution of a current stock of population, it is best to to account for population subgroups
with different mortality and to account for uncertainty about future changes in
mortality.
The present inquiry approximates thanatological age using lifetable-projected remaining lifetime,
which is intended to approximate remaining lifetime under the strong
assumptions of homogenous populations und fixed vital rates. 
%Chronological age-classified counts are redistributed over future death times
% according to a fixed mortality schedule.
%In general, a lifetable approach to transforming chronologically classified
%data into thanatologically classified data is sound if the given lifetable
%accurately describes the future mortality of the age-classified population, and
%if the chronological and thantological age patterns are independent.
In this case, each birth cohort undergoes the same
pattern of attrition. In a period setting, this means that we can distribute the
remaining lifespans of the population in each age class using the same
lifetable.
%This is the case we explore, and so the following formulas assume a single lifetable pattern.

@@

A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
$d_a$ column of the lifetable (here with single ages):
\begin{equation}
\label{eq:dyadiscret}
f_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \ec
\end{equation}
where $L_a$ is lifetable exposure.\footnote{Equation~\eqref{eq:dyadiscret}
will incur some error depending on the method used to derive $L_a$. To remove
this error, rescale over the $a$ margin so that each distribution conditioned on
$a$ sums to 1. Working with $\ell_a$ rather than $L_a$ eliminates error and
yields similar results, but is not necessarily better.} One produces in this way a probability density function over future death times for each chronological age,
$a$. The $f(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using\footnote{An
different-looking but identical definition is found in
\citet{brouard1989mouvements}.}

@@

Since $1 = \int_0^\infty f(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age. For
example, under these assumptions one can break down a population pyramid into
thanatological age classes.
Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data, assuming the
2010 period mortality schedule remains constant. When looking from the
chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this recomposition for the same 2010
US data.
